Highly Available Transactions: Virtues and Limitations

### 2013-04-25 writeup/intro.tex
adam's comments

The last decade has seen a large shift in the design of large-scale
database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. A prominent event in this shift
was the 2000 introduction of Brewer's CAP Theorem, which stated that a
highly available system cannot provide strong consistency guarantees
in the presence of network partitions~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concerns a fairly narrow form of
distributed consistency---are often conflated with the ability to
provide ACID database properties~\cite{brewer-slides, hn,
  foundation-article}; this has led to substantial confusion regarding
replica consistency, transactional isolation, and high
availability. While transactions \textit{are} useful for programmers
as evidenced by the recent resurgence of transactional data stores
such as Spanner~\cite{spanner}, these systems are unfortunately not
highly available in the presence of partitions~\cite{middleware-db,
  foundation-article, hstore,generalizedsnapshot, mdcc,
  krikellas-bargain, eiger, walter, calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable with high availability in the
presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-virtues,
  gray-isolation, kemme-thesis}. Due to their concurrency and overall
performance benefits, weak isolation models are widely employed,
overwhelmingly as the default setting in today's traditional ``ACID''
and up-and-coming ``NewSQL'' databases and often as the maximum
guarantee offered (Section~\ref{sec:modernacid}). While weak isolation
levels do not provide serializability for general-purpose
transactions, they are apparently strong enough to deliver acceptable
behavior to many application programmers and are substantially
stronger than the semantics provided by current highly available
systems. This raises a natural question: which of these alternative
semantics can be provided while remaining highly available?

Unfortunately, the relationship of high availability and ACID
semantics has not been well explored. We have a strong understanding
of weak isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. Additionally, the distributed computing, programming
language, and parallel hardware literature contains many
consistency models for single operations on replicated objects~\cite{
  ceri-mechanism, chen-mechanism, pnuts, herlihy-art, eiger, cac,
  sessionguarantees}. However, the literature has largely ignored
semantic guarantees for multiple operations operating on multiple data
items in a highly available distributed environment.

Our main contributions in this paper are as follows. We fill the gap
between, on one hand, the plethora of previously proposed isolation
and consistency models, and, on the other hand, the goal of high
availability.  We classify which among the wide array of models are
achievable with high availability, denoting them as {\em Highly
  Available Transactions} (HATs). In doing so, we demonstrate that many
implementations of HAT semantics are not highly available but
that their unavailability is not inherently required. Our
investigation shows, somewhat surprisingly, that besides
serializability, only Snapshot Isolation and Repeatable Reads are not
HAT-compliant, while most other isolation levels are achievable with
high availability. We also show that many weak data consistency models
from distributed systems are also HAT-compliant.

Making HAT classifications requires study of several phenomena. For
example, Snapshot Isolation (SI) and Repeatable Reads (RR) are not
HAT-compliant because they require detecting conflicts between
concurrent updates (as needed for preventing Lost Updates or Write
Skew phenomena). However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is easily implementable
and widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We study the virtues
and limitations of both classic and HAT systems by surveying
practitioner accounts and research literature, performing experimental
analysis on modern cloud infrastructure, and analyzing representative
applications for their semantic requirements. While our results are
not definitive, they suggest that HATs offer a one to three order of
magnitude latency decrease compared to traditional distributed
serializability protocols, and they can provide acceptable semantics
for a wide range of programs, especially those with monotonic logic
and commutative updates~\cite{calm, blooml, crdt}. HAT systems can
also enforce arbitrary foreign key constraints on multi-item updates
and sometimes provide limited uniqueness guarantees. However,
particularly for programs with non-monotonic logic, HATs can fall
short, requiring sometimes-unavailable mechanisms that are more
coordination intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.

>>>>

The last decade has seen a significant shift in the design of
large-scale database backends, from the use of transactional
RDBMSs~\cite{bernstein-book, gray-isolation, gray-virtues} to the
widespread adoption of loosely-consistent distributed key-value
stores~\cite{bigtable, pnuts, dynamo}. The 2000 introduction of
Brewer's CAP Theorem, which stated that a highly available system
cannot provide strong consistency guarantees in the presence of
network partitions, helped spur this shift~\cite{brewer-slides}. As
formally proven~\cite{gilbert-cap}, the CAP Theorem concerns
linearizability: the ability to read the most recent write to a data
item that is replicated across servers~\cite{herlihy-art}. Yet the
implications of CAP---which concern a fairly narrow form of
distributed consistency---have often been conflated with the ability
to provide ACID database properties~\cite{brewer-slides, hn,
  foundation-article}; this misunderstanding has led to substantial
confusion regarding replica consistency, transactional isolation, and
high availability. The recent resurgence of transactional systems like
Spanner~\cite{spanner} suggests that programmers value these
semantics, but, unfortunately, existing transactional data stores do
not provide availability in the presence of
partitions~\cite{middleware-db, foundation-article,
  hstore,generalizedsnapshot, mdcc, krikellas-bargain, eiger, walter,
  calvin}.

Indeed, serializable transactions---the gold standard of traditional
ACID databases---are not achievable with high availability in the
presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-virtues,
  gray-isolation, kemme-thesis}. Today's ACID and NewSQL databases
often employ weak isolation models due to their concurrency and
performance benefits; weak isolation is overwhelmingly the default
setting in these stores and is often the only option offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics provided by current highly available systems. This raises a
natural question: which of these semantics can be provided while
remaining highly available?

Unfortunately, the relationship between high availability and ACID
semantics has not been well-explored. We have a strong understanding
of weak isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-book, spanner, daudjee-session,
  hstore, krikellas-bargain, calvin, kemme-classification} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. Additionally, the distributed computing, programming
languages, and parallel hardware literature contains many
consistency models for single operations on replicated objects~\cite{
  ceri-mechanism, chen-mechanism, pnuts, herlihy-art, eiger, cac,
  sessionguarantees}. However, the literature has largely ignored
semantic guarantees for multiple operations operating on multiple data
items in a highly available distributed environment.

Our main contributions in this paper are as follows. We fill the gap
between the plethora of previously proposed isolation and consistency
models and the goal of high availability.  We classify which among the
wide array of models are achievable with high availability, denoting
them as {\em Highly Available Transactions} (HATs). In doing so, we
demonstrate that although many implementations of HAT semantics are
not highly available, this is an artifact of the implementations
rather than an inherent property of the semantics. Our investigation
shows that, besides serializability, only Snapshot Isolation and
Repeatable Reads are not HAT-compliant, while most other isolation
levels are achievable with high availability. We also demonstrate that
many weak data consistency models from distributed systems are also
HAT-compliant.

Our investigation is based on both impossibility results and several
constructive proof-of-concept algorithms. For example, Snapshot
Isolation (SI) and Repeatable Reads (RR) are not HAT-compliant because
they require detecting conflicts between concurrent updates (as needed
for preventing Lost Updates or Write Skew phenomena), which we prove
is unavailable. However, Read Committed isolation, Transactional
Atomicity, and many of the weaker data consistency models from
database and distributed systems are achievable via novel algorithms
that rely on multi-versioning and limited client-side caching. For
several guarantees, such as causal consistency with phantom prevention
and ANSI Repeatable Read, we consider a modified form of high
availability in which clients ``stick to'' (i.e., have affinity with)
at least one server (a property which is often implicit in the
distributed systems literature~\cite{herlihy-art, eiger, cac} but
which requires explicit consideration in a client-server replicated
database context). This sticky availability is easily implementable
and widely employed~\cite{eiger, vogels-defs} but is strictly less
available than traditional definitions of high availability.

Our results demonstrate that a wide range of semantic guarantees can
be delivered with availability in face of partitions, but which
guarantees are worthwhile in practice? We study the virtues and
limitations of both classic and HAT systems by surveying practitioner
accounts and research literature, performing experimental analysis on
modern cloud infrastructure, and analyzing representative applications
for their semantic requirements. Our experiences with a HAT prototype
running across multiple geo-replicated datacenters indicate that HATs
offer a one to three order of magnitude latency decrease compared to
traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs can fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits of
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.
